Reproducing anaylses of Beninca et al (2008)

### 2016-03-29 code/report.Rmd
various revisions

---
title: "Reproduce anaylses of Beninca et al (2008)"
author: "Owen Petchey et al"
date: "29 Jan 2015"
output:
  html_document:
    keep_md: yes
    toc: yes
  pdf_document:
    toc: yes
---

@@

This reproduction was made as part of the Reproducible Research in Ecology, Evolution, Behaviour, and Environmental Studies (RREEBES) Course, lead by Owen Petchey at the University of Zurich. More information about the course [here](https://github.com/opetchey/RREEBES/blob/master/README.md) on github.

The code and data for the reproduction are [here](https://github.com/opetchey/RREEBES/tree/Beninca_development) on github.

@@

@@

The Protozoa variable contains some numbers with comman as the decimal separator. This creates a question about what dataset was used for the original analyses, as it could not have been this one.
```{r}
spp.abund$Protozoa <- as.numeric(str_replace(spp.abund$Protozoa, ",", "."))
```

Format the dates as dates.
```{r}
spp.abund$Date <- dmy(spp.abund$Date)
# y123 <- year(spp.abund$Date) %% 100
# year(spp.abund$Date) <- 1900 + y123
# rm(y123)
```

Check dates match the Day.number (should give true):
```{r}
sum(spp.abund$Day.number == 1+as.numeric((spp.abund$Date - spp.abund$Date[1]) / 24 / 60 / 60)) == length(spp.abund$Date)
```

@@

Data is in wide format, so change it to long:
```{r}
spp.abund <- gather(spp.abund, "variable", "value", 3:12)
str(spp.abund)
```

@@

nuts <- select(nuts, -X, -X.1)
nuts <- nuts[-349:-8163,]
nuts$Date <- dmy(nuts$Date)
nuts <- select(nuts, -NO2, -NO3, -NH4)
#nuts$Date[duplicated(nuts$Date)]
#which(duplicated(nuts$Date))
nuts <- gather(nuts, "variable", "value", 3:4)
str(nuts)
```

@@

<!--
Here are some versions of the graphs that are not included in the report, hence the html comment:

@@

Make a sequence of times at which to interpolate. 
```{r}
#aggregate(Day.number ~ variable, all.data, min)
#aggregate(Day.number ~ variable, all.data, max)
#xout <- seq(343.35, 2657.2, by=3.35)
xout <- seq(343.35, 2658, by=3.35)
#range(xout)
```

@@

# run the function and tidy the data
# (somewhat ashamed of this code...)
# remove NAs before interpolation, so they don't cause problems
trimmed_raw <- na.omit(trimmed_raw)
res <- group_by(trimmed_raw, variable) %>%
  do(le=Interpolate(.$Day.number, .$value))

@@

# for(p in 1:nrow(res)){
#     temp <- data.frame(variable=res[p,1],
#                        res[p,2][[1]])
#     if(p==1)
#       trimmed_interp <- temp
#     if(p>1)
#       trimmed_interp <- rbind(trimmed_interp, temp)
#   }
```

@@

```{r}
## standardise
scaled_trimmed_final <- group_by(detrended, variable) %>%
  mutate(stand.y=as.numeric(scale(dt.value)))
#summarise(scaled_trimmed_final, mean=mean(stand.y), sd=sd(stand.y))
```

@@

# ggplot(filter(two.datasets), aes(x=as.numeric(Day.number), y=stand.y)) +
#   geom_path() +
#   facet_wrap(~variable) +
#   geom_line(aes(y=Abundance), col="red")

@@

@@

# freq.est <- spec %>% group_by(group) %>% mutate(max_spec = max(spec), freq = freq)
# freq.est <- subset(freq.est, max_spec==spec, select=c(freq,group))
# freq.est$freq <- 1/freq.est$freq

@@

# freq.est <- wspec %>%
#   group_by(group) %>%
#   mutate(max_spec = max(spec), freq = freq)
# freq.est <- subset(freq.est, max_spec==spec, select=c(freq,group))
# freq.est$freq <- 1/freq.est$freq
# #frequency(final$y)
# ts <- as.ts(final$y, frequency = 0.3)
# #time(ts)

@@

Get p-vals:
```{r}
cor.pvals <- matrix(NA, length(final_wide[1,]), length(final_wide[1,]))
for(i in 1:length(final_wide[1,]))
  for(j in 1:length(final_wide[1,])) {
    if(sum(!is.na(final_wide[,i]) & !is.na(final_wide[,j]))>10)
    cor.pvals[i,j] <- cor.test(final_wide[,i], final_wide[,j])$p.value
}
    #plot(final_wide[,2], final_wide[,3])
```

@@

Cosmetic work on the table:
```{r}
dim(cor.coefs) ## should be 12,12
oo <- c(10, 9, 8, 11, 12, 6, 5, 3, 4, 2)
cor.coefs1 <- cor.coefs[oo,oo]
cor.cp1 <- cor.cp[oo,oo]
for(i in 1:length(cor.cp1[1,]))
  for(j in 1:i)
  cor.cp1[i,j] <- ""
#sn <- c("Bacteria", "Harps", "Ostr", "N", "P", "Picophyt", "Nanophyt", "Rotifers", "Protozoa", "Calanoids")
sn <- substr(rownames(cor.cp1), 1, 4)
sn[4:5] <- c("N", "P")
dimnames(cor.cp1) <- list(sn, sn)
```

@@

Time delay (1), embedding dimension (6), and Theiler window (50) were used in the Nature report. Other parameters are chosen rather randomly!

@@

Do the time-delayed embedding, and save the data for the graphs:
```{r, message=FALSE, error=TRUE}
all.species <- unique(as.character(led$variable))
diverg <- matrix(NA, time.steps, length(all.species))
colnames(diverg) <- all.species
for(i in 1:length(all.species)) {
  print(all.species[i])
  tr.fs <- filter(final, variable==all.species[i])$y
  diverg[,i] <- as.numeric(try(lyap_k(tr.fs,
                                      m=embedding.dimension,
                                      d=time.delay,
                                      k=10, # number of considered neighbours 20
                                      ref=40, # number of points to take into account 100
                                      t=Theiler.window,
                                      s=time.steps,
                                      eps=10 # radius where to find nearest neighbours 10
                                      )))
}
## a bit of a fudge with the translation to days
diverg <- as.data.frame(cbind(days=1:time.steps, diverg))
diverg <- gather(diverg, Species, Difference, 2:10)
diverg$days <- diverg$days*3.35
#str(diverg)
```

Next calculate the Lyapunov exponents, noting that 6 or 7 points were used in the regressions in the Nature report
```{r}
#diverg$Difference[is.na(diverg$Difference)] <- 0
diverg$Difference[is.infinite(diverg$Difference)] <- NA
diverg.short <- filter(diverg, days<24) ## 24 is about 6 steps, after initial gap
LEs <- group_by(diverg.short, Species) %>%
  summarise(le=coef(lm(Difference[1:6] ~ days[1:6]))[2])
#pval=summary(lm(Difference[1:6] ~ days[1:6]))$coefficients[2,4])
```

Then plot the graphs with LE:
```{r}
LEs <- mutate(LEs, days=20, Difference=-0.5)
diverg$Species_short <- plyr::revalue(diverg$Species,
                                 c(Soluble.reactive.phosphorus="P",
                                   Total.dissolved.inorganic.nitrogen="N"))
LEs$Species_short <- plyr::revalue(LEs$Species,
                                 c(Soluble.reactive.phosphorus="P",
                                   Total.dissolved.inorganic.nitrogen="N"))
g1 <- ggplot(diverg, aes(x=days, y=Difference)) + geom_point() + facet_wrap(~Species_short) +
  geom_text(data=LEs, aes(x=60, y=0, label=round(le,3)))
g1
```

@@

## same but to allow x-y scatter
mt$Day.number <- as.character(mt$Day.number)
from.steve$Day.number <- as.character(from.steve$Day.number)
ff <- inner_join(mt, from.steve)
mt$Day.number <- as.numeric(mt$Day.number)
from.steve$Day.number <- as.numeric(from.steve$Day.number)
ff$Day.number <- as.numeric(ff$Day.number)
#str(ff)
#ff$difference <- ff$value - ff$value_steve
#qplot(log10(ff$difference))
qplot(data=ff, x=log10(value_steve), y=log10(value)) +
  xlab("Values from Ellner\n(log 10 transformed)") +
  ylab("Reproduced values\n(log 10 transformed)") +
  geom_abline(intercept=0, slope=1, colour="red")

@@

```{r}
## standardise
#scaled_trimmed_final <- group_by(trimmed_detrended, variable) %>%
#  mutate(stand.y=as.numeric(scale(dt.value)))
#summarise(scaled_trimmed_final, mean=mean(stand.y), sd=sd(stand.y))

## or don't standardise
final <- detrended
final$y <- final$dt.value
#summarise(final, mean=mean(y), sd=sd(y))
```

@@

```{r}
## use next line to work on unstandardised data
final.to.melt <- final[, c("variable", "dt.value", "Day.number")]
## use next line to work on standardised
#final.to.melt <- final[, c("variable", "y", "Day.number")]
names(final.to.melt)[1] <- "Species"
melted <- melt(final.to.melt, id=c("Species", "Day.number"))
X <- acast(melted, formula= Day.number ~ Species)
#str(X)
X <- as.data.frame(X)
```

@@

cors <- data.frame(dist=1:12, rsq_vals)
cors.long <- gather(cors, key=dist)
names(cors.long) <- c("Prediction_distance", "Variable", "Correlation")

@@

## read in the data of the original article that came direct from Elisa
original_preds <- read.csv(text=getURL(paste0(repo,"/data/original/original_rsquared.csv")), skip=0, header=T)
opw <- gather(original_preds, Species, r_squared, 2:25)
opw <- separate(opw, Species, c("Species", "Model"))

name.mapping <- data.frame(Species=c("cyclo", "roti", "cala", "pico", "nano",
                                   "dia", "P", "N", "bact", "ostra", "harpa",
                                   "proto"),
                           Variable=c("Cyclopoids", "Rotifers", "Calanoid.copepods",
                                  "Picophytoplankton", "Nanophytoplankton",
                                  "Filamentous.diatoms", "Soluble.reactive.phosphorus",
                                  "Total.dissolved.inorganic.nitrogen", "Bacteria",
                                  "Ostracods", "Harpacticoids", "Protozoa"))
opw <- full_join(opw, name.mapping)

>>>>

---
title: "Reproducing anaylses of Beninca et al (2008)"
author: "Owen Petchey, Marco Plebani, Frank Pennekamp"
date: "28 Mar 2015"
output:
  html_document:
    keep_md: yes
    toc: yes
  pdf_document:
    toc: yes
---

@@

This reproduction was made as part of the Reproducible Research in Ecology, Evolution, Behaviour, and Environmental Studies (RREEBES) Course, lead by Owen Petchey at the University of Zurich. More information about the [RREEBES course on github](https://github.com/opetchey/RREEBES/blob/master/README.md) .

[The code and data for the reproduction are on github.](https://github.com/opetchey/RREEBES/tree/Beninca_development).

@@

# Testing

This code in this document has been tested on:

```
platform       x86_64-apple-darwin14.5.0   
arch           x86_64                      
os             darwin14.5.0                
system         x86_64, darwin14.5.0        
status                                     
major          3                           
minor          2.3                         
year           2015                        
month          12                          
day            10                          
svn rev        69752                       
language       R                           
version.string R version 3.2.3 (2015-12-10)
nickname       Wooden Christmas-Tree   
```

and

```
Win 7 64-bit with R 3.1.3.
```

@@

The Protozoa variable contains some numbers with comma as the decimal separator. This creates a question about what dataset was used for the original analyses, as it could not have been this one.
```{r}
spp.abund$Protozoa <- as.numeric(str_replace(spp.abund$Protozoa, ",", "."))
```

Format the dates as dates.
```{r}
spp.abund$Date <- dmy(spp.abund$Date)
```

Check dates match the Day.number (should give TRUE):
```{r}
stopifnot(sum(spp.abund$Day.number == 1+as.numeric((spp.abund$Date - spp.abund$Date[1]) / 24 / 60 / 60)) == length(spp.abund$Date))
```

@@

Data is in wide format, so change it to long:
```{r}
spp.abund <- gather(spp.abund, key="variable", value="value",
                    3:12)
spp.abund$variable <- as.factor(spp.abund$variable)
str(spp.abund)
```

@@

nuts <- select(nuts, -X, -X.1)
nuts <- nuts[-349:-8163,]
nuts$Date <- dmy(nuts$Date)
nuts <- select(nuts, -NO2, -NO3, -NH4)
nuts <- gather(nuts, "variable", "value", 3:4)
nuts$variable <- as.factor(nuts$variable)
str(nuts)
```

@@

<!--
Here are some versions of the graphs that were interesting to produce originally, and may be of some interest, but are not included by default for brevity:

@@

Make a sequence of times at which to interpolate. 
```{r}
xout <- seq(343.35, 2658, by=3.35)
```

@@

# run the function and tidy the data
# remove NAs before interpolation, so they don't cause problems
trimmed_raw <- na.omit(trimmed_raw)
res <- group_by(trimmed_raw, variable) %>%
  do(le=Interpolate(.$Day.number, .$value))

@@

@@

```{r}
## standardise
scaled_trimmed_final <- group_by(detrended, variable) %>%
  mutate(stand.y=as.numeric(scale(dt.value)))
```

@@

## some graphs that are of use for very detailed views of the two datasets. Not included in report by default for brevity.
# ggplot(filter(two.datasets), aes(x=as.numeric(Day.number), y=stand.y)) +
#   geom_path() +
#   facet_wrap(~variable) +
#   geom_line(aes(y=Abundance), col="red")

@@

# ensure alphabetical order of panels in figure
two.datasets$variable <- factor(two.datasets$variable,
                                 levels = sort(levels(as.factor(two.datasets$variable))))

@@

@@

@@

Get p-vals:
```{r}
cor.pvals <- matrix(NA, length(final_wide[1,]), length(final_wide[1,]))
for(i in 1:length(final_wide[1,]))
  for(j in 1:length(final_wide[1,])) {
    if(sum(!is.na(final_wide[,i]) & !is.na(final_wide[,j]))>10)
    cor.pvals[i,j] <- cor.test(final_wide[,i], final_wide[,j])$p.value
}
```

@@

Cosmetic work on the table:
```{r}
oo <- c(10, 9, 8, 11, 12, 6, 5, 3, 4, 2)
cor.coefs1 <- cor.coefs[oo,oo]
cor.cp1 <- cor.cp[oo,oo]
for(i in 1:length(cor.cp1[1,]))
  for(j in 1:i)
  cor.cp1[i,j] <- ""
sn <- substr(rownames(cor.cp1), 1, 4)
sn[4:5] <- c("N", "P")
dimnames(cor.cp1) <- list(sn, sn)
```

@@

Time delay (1), embedding dimension (6), and Theiler window (50) were used in the Nature report. Other parameters are chosen rather arbitrarily.

@@

Do the time-delayed embedding, and save the data for the graphs:
```{r, message=FALSE, error=TRUE}
all.species <- unique(as.character(led$variable))
diverg <- matrix(NA, time.steps, length(all.species))
colnames(diverg) <- all.species
for(i in 1:length(all.species)) {
  print(all.species[i])
  tr.fs <- filter(final, variable==all.species[i])$y
  diverg[,i] <- as.numeric(try(lyap_k(tr.fs,
                                      m=embedding.dimension,
                                      d=time.delay,
                                      k=10, # number of considered neighbours 20
                                      ref=40, # number of points to take into account 100
                                      t=Theiler.window,
                                      s=time.steps,
                                      eps=10 # radius where to find nearest neighbours 10
                                      )))
}
## a bit of a fudge with the translation to days
diverg <- as.data.frame(cbind(days=1:time.steps, diverg))
diverg <- gather(diverg, Species, Difference, 2:10)
diverg$days <- diverg$days*3.35
```

Next calculate the Lyapunov exponents, noting that 6 or 7 points were used in the regressions in the Nature report
```{r}
diverg$Difference[is.infinite(diverg$Difference)] <- NA
diverg.short <- filter(diverg, days<24) ## 24 is about 6 steps, after initial gap
LEs <- group_by(diverg.short, Species) %>%
  summarise(le=coef(lm(Difference[1:6] ~ days[1:6]))[2])
```

Then plot the graphs with LE:
```{r}
LEs <- mutate(LEs, days=20, Difference=-0.5)
diverg$Species_short <- plyr::revalue(diverg$Species,
                                 c(Soluble.reactive.phosphorus="P",
                                   Total.dissolved.inorganic.nitrogen="N"))
LEs$Species_short <- plyr::revalue(LEs$Species,
                                 c(Soluble.reactive.phosphorus="P",
                                   Total.dissolved.inorganic.nitrogen="N"))

## ensure order of panels in figure same as in original
diverg$Species_short <- factor(diverg$Species_short,
                               levels=c("Calanoid.copepods", "Rotifers",
                                        "Nanophytoplankton", "Picophytoplankton",
                                        "Ostracods", "Harpacticoids", "Bacteria",
                                        "N","P"))

g1 <- ggplot(diverg, aes(x=days, y=Difference)) + geom_point() + facet_wrap(~Species_short) +
  geom_text(data=LEs, aes(x=60, y=0, label=round(le,3)))
g1
```

@@

## same but to allow x-y scatter
mt$Day.number <- as.character(mt$Day.number)
from.steve$Day.number <- as.character(from.steve$Day.number)
ff <- inner_join(mt, from.steve)
mt$Day.number <- as.numeric(mt$Day.number)
from.steve$Day.number <- as.numeric(from.steve$Day.number)
ff$Day.number <- as.numeric(ff$Day.number)
qplot(data=ff, x=log10(value_steve), y=log10(value)) +
  xlab("Values from Ellner\n(log 10 transformed)") +
  ylab("Reproduced values\n(log 10 transformed)") +
  geom_abline(intercept=0, slope=1, colour="red")

@@

```{r}
## standardise (commented out by default; uncomment, and comment next chunck in order to work on standardised data)
#scaled_trimmed_final <- group_by(trimmed_detrended, variable) %>%
#  mutate(stand.y=as.numeric(scale(dt.value)))
#summarise(scaled_trimmed_final, mean=mean(stand.y), sd=sd(stand.y))
```

```{r}
## or don't standardise
final <- detrended
final$y <- final$dt.value
```

@@

```{r}
## use next line to work on unstandardised data
final.to.melt <- final[, c("variable", "dt.value", "Day.number")]
## use next line to work on standardised
names(final.to.melt)[1] <- "Species"
melted <- melt(final.to.melt, id=c("Species", "Day.number"))
X <- acast(melted, formula= Day.number ~ Species)
X <- as.data.frame(X)
```

@@

cors <- data.frame(dist=1:12, rsq_vals)
cors.long <- gather(cors, key=dist)
names(cors.long) <- c("Prediction_distance", "Variable", "Correlation")
cors.long$Variable <- as.factor(cors.long$Variable)

@@

## read in the data of the original article that came direct from Elisa
original_preds <- read.csv(text=getURL(paste0(repo,"/data/original/original_rsquared.csv")), skip=0, header=T)
opw <- gather(original_preds, Species, r_squared, 2:25)
opw <- separate(opw, Species, c("Species", "Model"))
opw$Species <- as.factor(opw$Species)
opw$Model <- as.factor(opw$Model)
name.mapping <- data.frame(Species=c("cyclo", "roti", "cala", "pico", "nano",
                                   "dia", "P", "N", "bact", "ostra", "harpa",
                                   "proto"),
                           Variable=c("Cyclopoids", "Rotifers", "Calanoid.copepods",
                                  "Picophytoplankton", "Nanophytoplankton",
                                  "Filamentous.diatoms", "Soluble.reactive.phosphorus",
                                  "Total.dissolved.inorganic.nitrogen", "Bacteria",
                                  "Ostracods", "Harpacticoids", "Protozoa"))
opw <- full_join(opw, name.mapping)
